  \documentclass[11pt,fleqn]{article}
 \usepackage{amsmath}
 \usepackage{amssymb}
 \usepackage{bm}
 \usepackage[margin=1in]{geometry}
 \usepackage{multicol}
\usepackage{fancyhdr}
\usepackage{systeme}
\usepackage{enumerate}
\usepackage{nicefrac}

\pagestyle{fancy}
\fancyfoot{}
\lhead{{\bf LinAlg Math Camp}}
\rhead{{\bf Worksheet \#5}}
\chead{}
\cfoot{\thepage}

\newcommand{\mat}[1]{\mathbf{#1}}
\newcommand{\imag}{\mathrm{i}}

% Actual document starts here 
% ======================================================================================
\begin{document}

\begin{enumerate}
\item Let $\mathbf{A}$ and $\mathbf{B}$ both be $n\times n$ and assume that there is an invertible matrix $\mathbf{S}$ such that $\mathbf{SAS}^{-1}$ and  $\mathbf{SBS}^{-1}$ are both diagonal (but not necessarily equal). Prove that $\mathbf{AB} = \mathbf{BA}$.

\item Suppose that $\mathbf{A}$ is symmetric and positive definite, while $\mathbf{B}$ is symmetric. Prove that there is a real invertible matrix $\mathbf{S}$ such that
\[\mathbf{A} + \mathbf{B} = \mathbf{S}\left(\mathbf{I} + \mathbf{D}\right)\mathbf{S}^T\]
where $\mathbf{D}$ is diagonal.

\item Suppose that $\mathbf{A}$ and $\mathbf{B}$ are both $n\times n$. Prove that $\mathbf{AB}$ and $\mathbf{BA}$ have the same eigenvalues. % 8.2.23

\item Suppose that $\mathbf{L}\in\mathbb{R}^{n\times n}$ is a symmetric matrix with eigenvalues $\lambda_1,\ldots,\lambda_m$. (Note that $m$ can be $\le n$.) Prove that
\[\mathbf{L} = \lambda_1 \mathbf{P}_1 + \cdots+\lambda_m\mathbf{P}_m\]

\item Prove that if $\mathbf{A}$ is invertible (and, for simplicity, real), then there is an orthogonal matrix $\mathbf{U}$ and a symmetric and positive definite matrix $\mathbf{E}$ such that $\mathbf{A} = \mathbf{UE}$. This is analogous to writing complex numbers in polar form $x+\imag y = \rho e^{\imag \theta}$. Show that if $\mathbf{A}$ is square but not invertible then there is a similar factorization but with $\mathbf{E}$ being symmetric and non-negative definite.

\item Let $\mathbf{A} = \mathbf{S\Lambda S}^{-1}$ be diagonalizable. Find a symmetric positive definite matrix $\mathbf{K}$ that defines an inner product $\langle\vec{x},\vec{y}\rangle =\vec{x}^T\mat{K}\vec{y}$ such that $\mathbf{A}$ commutes with its $\mat{K}$-adjoint.

\item Let $\mat{A}\in\mathbb{R}^{m\times n}$ have full column rank. Prove that $\|\mat{A}(\mat{A}^T\mat{A})^{-1}\|_2=1/\sigma_n$ where $\sigma_n$ is the $n^{\mbox{\tiny th}}$ singular value of $\mat{A}$. (Note that $\mat{A}^{-1}$ is not defined, so your proof cannot use $\mat{A}^{-1}$.)

where $\mathbf{P}_i$ is an orthogonal projection matrix that projects orthogonally onto the nullspace of $\mathbf{L} - \lambda_i\mathbf{I}$.
\end{enumerate}
 

\end{document}